Contracts for unions and intersections are hard for Nickel

### 2021-05-16 paper.tex
Review on other papers and Racket

\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}

@@

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together.
They provide two different set of rules for their calculus,
the first one is non deterministic, which purpose is to provide
semantics that are simple to understand.
The second one is a deterministic set of rules, that are intended
to be simply to implement. These semantics need to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.

@@

$$ K \in \llbracket A \cap B \rrbracket ~ if ~ K \in \llbracket A \rrbracket \lor K \in \llbracket B \rrbracket $$

Which means that a continuation ($K$) complies negatively with an
intersection of $A$ and $B$ if it complies with at least one of these.
Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
where $K_1 \in \llbracket A \rrbracket$ but $K_1 \not \in \llbracket B \rrbracket$,
and dually for $K_2$, which would mean the continuation presented above would not
be in $\llbracket A \cap B \rrbracket$, even though our intuition (and the
introduction to the paper) indicates that it should.

While this mistake does not invalidate their work (the system they present
handles these correctly), it shows that existing research is still not on
good shape to handle the growing usecases being developed for
union and intersection around dynamic languages.

@@

% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
\unsure{Is this worth mentioning? Should we go so techincal?}

@@

Unions and intersection are starting to become more and more common
on non-academic programming languages, as stated on the introduction,
unions are a natural abstraction for dynamic languages, where the type
of an expression could have many different shapes, without necessarily
having a mark to distinguish between them.
With this in mind, a valid question is how do these languages solve the
union/intersection conundrum? Or, even better, what is it that they
solve?

@@

Racket is language based on Scheme mainly thought as a language to experiment
with different language desings and ideas. It provides a very complete contract
system.

Among those contracts, they provide \texttt{and/c} and \texttt{or/c} contracts.
These contracts are lightweight versions of unions and intersections, they simply
check that every contract holds, in the case of and/c, and that at least one contract
holds, in the case of or/c.\unsure{It doesn't really make sense for a
  contract to “hold”, in general}

So, for instance, the usual usecase of intersection to encode overloading
does not work on Racket. Take a look at the next example, the function
\texttt{overload} should accept a call like \texttt{(overload 3)}, but it fails
since it 3 is not a valid \texttt{string?}.

\texttt{(define/contract overload
(and/c (-> number? number?)
       (-> string? string?))
($\lambda$ (x) x))}

Even worst, the or/c contract doesn't allow to have multiple higher order
contracts that could (potentially) apply to a function, these have to be
differentiated by first order markers (for instance, the number of
arguments).\unsure{This section is overall a bit hard to follow. I
  think I've figured out what it means, but it needs more love. Also
  give an example which doesn't work and one that does for or/c.}

@@

Racket does provide an alternative for overloading methods, called the
\texttt{case->} contract, however, this overloading of methods only
works over contracts of functions with different amount of parameters.

>>>>

\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=lisp]{#1}}

@@

For instance, if the value \nickel{V} is contracted with
\nickel{A /\\ (B \\/ C)}, then, before the contract is checked,
the contract gets opened up to something equivalent to
\nickel{(A /\\ B) \\/ (A /\\ C)}.

@@

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together.
\todo{This assumes the difficulty was already explained}
They provide two different sets of rules for their calculus,
the first one is non deterministic, whose purpose is to provide
semantics that are simple to understand.
The second one is a deterministic set of rules, that are intended
to be simple to implement.
This second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.

@@

$$ K \in \llbracket A \cap B \rrbracket^- ~ if ~ K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$

Which means that a continuation ($K$) complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.
Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
and dually for $K_2$, which would mean the continuation presented above would not
be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
introduction to the paper) indicates that it should.

While this mistake does not invalidate their work (the system they present
handles these cases correctly), it shows that existing research is still not on
good shape to handle the growing usecases being developed for
union and intersection around dynamic languages.

@@

% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
\unsure{Is it worth mentioning how continuations are formed?
Should we go so techincal?}

@@

Unions and intersection are starting to become more and more common
on non-academic programming languages, as stated on the introduction,
unions are a natural abstraction for dynamic languages, where the type
of an expression could have many different shapes, without necessarily
having a mark to distinguish between them at type level.
With this in mind, a valid question is how do these languages solve the
union/intersection conundrum? Or, even better, what is it that they
solve?

@@

Racket is a language based on Scheme, mainly thought as a language to experiment
with different language desings and ideas. It provides a very complete contract
system\cite{RacketContracts}.

Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
These contracts are lightweight versions of intersections and unions, they simply
check that every contract is valid, in the case of \racket{and/c},
and that at least one contract
is valid, in the case of \racket{or/c}.

So, for instance, the usual usecase of intersection to encode bounded overloading
does not work on Racket. Take a look at the next example, the function
\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
since 3 is not a valid \racket{string?}.

\begin{figure}[h]

\begin{lstlisting}[language=lisp]
(define/contract overload
(and/c (-> number? number?)
       (-> string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{Non working function overloading on Racket.}

\end{figure}

Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
contracts that could (potentially) apply to a function, these have to be
differentiated by first order markers.
For instance, the code on Figure \ref{code:racket:or/c:working}
would correctly allow only to call
\racket{united} with a \racket{number?},
and it can make this distinction since it knows that
\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
However, the code on Figure \ref{code:racket:or/c:non-working} would fail
before running, since the \racket{or/c} contract is unable to know
which of the two function contracts to use.
Ideally, this last example should just behave as wrapping
the function with the \racket{(-> even? even?)} contract.

\begin{figure}[h]

\begin{lstlisting}[language=lisp]
(define/contract united
(or/c (-> number? number?)
      (-> string? string? string?))
(lambda (x) x))
\end{lstlisting}
\caption{Working \racket{or/c} example on Racket.}
\label{code:racket:or/c:working}

\end{figure}

\begin{figure}[h]

\begin{lstlisting}[language=lisp]
(define/contract united
(or/c (-> number? number?)
      (-> even? even?))
(lambda (x) x))
\end{lstlisting}
\caption{Non working \racket{or/c} example on Racket.}
\label{code:racket:or/c:non-working}

\end{figure}

\unsure{This section is overall a bit hard to follow. I
  think I've figured out what it means, but it needs more love. Also
  give an example which doesn't work and one that does for or/c.}

@@

Racket does provide an alternative for overloading methods, called the
\racket{case->} contract, however, this overloading of methods only
works over contracts of functions with different amount of parameters.
It's a good alternative to allow limited overloading of functions.
As an example, consider the following non failing piece of code, and
try to understand how Racket can decide, at each application,
which function contract to use.

\begin{figure}[h]

\begin{lstlisting}[language=lisp]
(define/contract overcase
  (case-> (-> string? string?)
          (-> number? number? number?)
          )
  (lambda (x [y 0]) (if (number? x)
                   (+ x y)
                   x)))

(overcase 1 2)

(overcase "hello")
\end{lstlisting}
\end{figure}
